Human capital mediates natural selection in contemporary humans

### 2022-03-26 response-to-reviewers-BE.Rmd

```

Dear Professor Keller,

Thank you very much for the chance to revise our manuscript. We would also like to thank both reviewers for their comments. Below we deal with the points they raised.

Code to produce this paper is now publicly available at <https://github.com/hughjonesd/why-natural-selection>.

@@

The within-family estimate should remove the worries about exogeneity of p\*, but they are estimated with a smaller sample. We agree that the estimate is high. The bootstrapped 95% confidence interval is from around 0% to 300%, which is not really informative, though it does suggest that genetic nurture is more likely to bias our results downwards than upwards. We've removed the discussion.

> p\. 40 and following: I confess to not following the derivations completely. The authors seem adept at what they do, and I trust that they have made no serious errors.
>
> (1) This is my first encounter with the Karush-Kun-Tucker conditions or the Inada condition.

KTK are the standard conditions for constrained optimization (see e.g. Sundaram 1996). We think people who want to follow the proofs will know about them. The Inada condition is a bit more specific to economics. We are now more explicit about what it means. It can be derived in this case by differentiating $u$ to give $u'(y) = y^{-\sigma}$ which for $\sigma > 0$ approaches infinity as $y \rightarrow 0$.

> (2) Below Equation 25, I did not verify the second derivative of N_2 with respect to h. This seems to require simple (if tedious) applications of the product and chain rules. The important conclusion, that N_2 is convex in h for sigma close to 1, is something that I will accept by analogy to the case N_1, N_2 \> 0.

@@

This is correct.

> (3) I do not understand in what way Equations 13 and 14 determine when the case N_1 = 0, N_2 \> 0 holds. Nevertheless, I accept the intuitive conclusion that it holds for intermediate values of h because too small a value of h (which determines s\*) will make N_1\* in Equation 17 greater than zero and because too high a value will make N_2\* in 18 negative. In a similar way, I accept that no children at all is the optimum for very high values of h; this seems reasonable for this model.

@@

> 7\. The authors write that sigma captures risk aversion in income in their model. But that's confusing and not a good way of describing this for their model, because as far as I understand there's no income risk! They should clarify this and say that sigma regulates the curvature of the utility function, and explain what this means in their riskless model.

It is a fair point. We have rewritten accordingly: "$\sigma > 0$ measures the curvature of the utility function, i.e. the decline in marginal utility of income as income increases".\
\
OTHER COMMENTS AND SUGGESTIONS

> 8\. Most figures are hard to read. For example, Figures 3-5 are hard to read. In Figure 3a, the dot color indicate education and the dot border color indicates significance; that's very difficult to read quickly (for instance, a blue dot with a yellow border looks green). Instead, the authors could use different symbols (crosses, squares) for education and colors for significance, or vice-versa.

We experimented a lot with our basic figures, and they may not be easy to improve. The graphs' key message is "effect sizes in one category are larger than others". To see that the eye needs to be able to quickly differentiate between categories. We have tried several variants using symbols, but in our judgment the added visual noise makes it harder to follow the categories across rows. The figure below shows three alternative versions: our original, a version using symbols, and one with subtle tweaks to colors and stroke sizes, which we hope fix the greenness issue and make p-values clearer. We prefer our tweaked version (and have used it in the revision), but we will bow to the reviewers or editor. For comparison, a version of the whole paper using the symbols plot is available [here](https://raw.githubusercontent.com/hughjonesd/why-natural-selection/alt-plots/why-natural-selection-weighted.pdf).

### 2022-03-28 why-natural-selection.Rmd
Remove formal analysis of effects on the income/EA3 correlation

cor_ea3 <- res_ineq[res_ineq$score == "EA3_excl_23andMe_UK", ]
cor_ea3$increase_pct <- (cor_ea3$ratio - 1) * 100

drake::loadd(res_ineq_ea3)

r2_income_true_psea <- scales::percent(res_ineq_ea3$r2_income_true_psea, 
                                         accuracy = 0.1)
r2_income_true_psea_wt <- scales::percent(res_ineq_ea3$r2_income_true_psea_wt, 
                                            accuracy = 0.1)
```

Because selection effects are concentrated in lower-income groups, they may also
increase inequality with respect to polygenic scores. For example, Figure
\@ref(fig:plot-mean-EA3-by-income) graphs mean polygenic scores for educational
attainment (EA3) among children from households of different income groups. The
blue bars show the actual means, i.e. parents' mean polygenic score weighted by
number of children. The grey bars show the hypothetical means if all households
had equal numbers of children. Natural selection against genes associated with
educational attainment, which lowers the mean, is stronger at the bottom of the
income distribution, and this increases the differences between groups.
Overall, natural selection increases the correlation of polygenic scores with
income for `r n_increases` out of 33 polygenic scores, with a median
percentage increase of `r median_pct_change`% (Appendix Table
\@ref(tab:tbl-inequality-effects)).

If inequalities in polygenic scores are important for understanding social
structure and mobility [@belsky2018genetic;
@Rimfeld_2018; @harden2021genetic], then these increases are substantive. A
back-of-the-envelope calculation, using estimated heritability for educational
attainment of 0.4 [@branigan2013variation], gives the proportion of variance in
household income attributable to the "true" polygenic score for educational
attainment as `r r2_income_true_psea` among respondents. Natural selection would
raise this proportion to `r r2_income_true_psea_wt` among respondents' children,
counterfactually holding environmental conditions constant. Similarly, since
many polygenic scores are predictive of disease risk, they could potentially
increase health inequalities. In general, the evolutionary history of
anatomically modern humans is related to disease risk [@benton2021influence];
understanding the role of contemporary natural selection may aid in mapping the
genetic architecture of current health disparities.

@@

## Estimating the effect of natural selection on the contribution of polygenic scores for educational attainment to inequality

```{r calc-load-ineq-ea3}

res <- drake::readd(res_ineq_ea3)

prop_effect_selection <- res$r2_income_ea3_wt/res$r2_income_ea3 - 1
prop_effect_selection <- scales::percent(prop_effect_selection, accuracy = 0.1)

prop_wt_fe <- res$r2_income_ea3_fe_wt/res$r2_income_ea3_fe - 1
prop_wt_fe <- scales::percent(prop_wt_fe, accuracy = 0.1)

prop_wt_fe_bounds <- scales::percent(res$r2_fe_bounds, accuracy = 0.1)

res <- lapply(res, format, digits = 3)

From the meta-analysis of twin studies in @branigan2013variation, we take the
heritability of educational attainment to be around 0.4. This is the variance of
educational attainment explained by the "true" polygenic score for educational
attainment (PSEA) $p^{*}$, i.e. the best linear unbiased predictor of educational
attainment from genetic data. We write the measured polygenic
score of educational attainment, "EA3" from @lee2018gene, as
\[
p = p^* + \eta.
\]

For realized educational attainment $EA$, write:
\[
EA = \alpha+\beta p^* + \varepsilon = \alpha + \beta(p - \eta) + \varepsilon
\]
We assume the classical errors-in-variables model, i.e. that $\eta$
is independent of both $p$ and $\varepsilon$. Under these conditions,
the ratio between the estimated $R^{2}$ in a regression of $p$ on
$EA$, and the true $R^{2}$ of $p^{*}$ on $EA$, satisfies: 
\begin{equation}
\frac{\textrm{plim}\hat{R}^{2}}{R^{2}} 
= \frac{\sigma_{p}^{2} - \sigma_{\eta}^{2}}{\sigma_{p}^{2}}
= \frac{\sigma_{p^*}^{2}}{\sigma_{p}^{2}}
\equiv \lambda, \label{eq:r2-ratio}
\end{equation}
as we show below. The right hand side is the "reliability ratio".
We regress educational attainment on EA3 in our sample, weighting
by our Age/Qualification weights, to give 
$\textrm{plim}\hat{R}^2 = `r res$r2_edu_ea3`$. Plugging this
into the above, with $R^2 = h^2\approx0.4$, gives $\lambda=`r res$lambda`$.

We now consider the $R^2$ of $p$ on income, that is, the proportion
of variance in income that is explained by the true polygenic score
of educational attainment. Applying (\ref{eq:r2-ratio}) again gives
that
\begin{equation}
\frac{\textrm{plim}\hat{R}_{\textrm{income}}^{2}}{R_{\textrm{income}}^{2}}=\lambda.\label{eq:r2-ratio-income}
\end{equation}

We compute $\hat{R}_{\textrm{income}}^{2}$by regressing income group on EA3,
first weighting by age/qualification, and then by age/qualification weights
times respondent's number of children. The estimated $R^2$ are 
`r res$r2_income_ea3` and `r res$r2_income_ea3_wt` respectively. Applying
(\ref{eq:r2-ratio-income}) to both estimates gives the proportion of variance in
income explained by true polygenic score for education attainment (a) among
respondents as `r res$r2_income_true_psea`; (b) accounting for natural selection
as `r res$r2_income_true_psea_wt`. The second figure gives the counterfactual
estimate of $R_{\textrm{income}}^{2}$ after one generation of natural selection,
holding environmental conditions unchanged.

The errors-in-variables conditions are quite restrictive. In particular, they
rule out that the unmeasured part of PSEA ($\eta$) correlates with the error 
term $\varepsilon$, e.g. with environmental causes of educational attainment. 
If unmeasured PSEA correlates positively with $\varepsilon$, then the
reliability coefficient $\lambda$ will be lower than calculated. However, the
proportional effect of natural selection on the PSEA-income relationship will
remain the same, so long as $\lambda$ does not differ between the two sets of
weighted regressions.

The estimate above uses a pooled regression of income on EA3, which may include
confounding environmental variation such as genetic nurture. Within-siblings
regression avoid this, but the smaller N makes estimates noisier. Running
within-siblings regressions of EA3 on income group gives within-$R^2$ estimates
of `r res$r2_income_ea3_fe` (weighted by age and qualification) and `r
res$r2_income_ea3_fe_wt` (multiplying weights by number of children). Thus, this
procedure estimates the proportional change in $R^2_{income}$ due to natural
selection as `r prop_wt_fe`. However, bootstrap 95% confidence intervals
are large (`r prop_wt_fe_bounds[1]` to `r prop_wt_fe_bounds[2]`). We therefore
view this result more as suggesting that environmental confounds likely
bias the estimated change in $R^2_{income}$ towards zero.

### Proof of \@ref(eq:r2-ratio)

Write 
\begin{align}
\hat{R}^{2} & =1-\frac{\sum_{i=1}^{N}(\hat{\alpha}+\hat{\beta}p_{i}-EA_{i})^{2}}{\sum_{i=1}^{N}(EA_{i}-\bar{EA})^{2}}\label{eq:r-squared}
\end{align}

Using the fact that $\hat{\alpha}=\bar{EA}-\hat{\beta}\bar{p}$, where bars signify
sample averages, we can simplify the numerator in the above:

\[
\sum_{i=1}^{N}(\hat{\alpha}+\hat{\beta}p_{i}-EA_{i})^{2}=\sum_{i=1}^{N}(\bar{EA}-\hat{\beta}\bar{p}+\hat{\beta}p_{i}-EA_{i})^{2}=\sum_{i=1}^{N}(\hat{\beta}(p_{i}-\bar{p})-(EA_{i}-\bar{EA}))^{2}
\]

Now, we can compute the following probability limit:

\begin{align}
 & \text{plim}\frac{\sum_{i=1}^{N}(\hat{\beta}(p_{i}-\bar{p})-(EA_{i}-\bar{EA}))^{2}}{N}\nonumber \\
= & \text{plim}\frac{\sum_{i=1}^{N}\hat{\beta}^{2}(p_{i}-\bar{p})^{2}+(EA_{i}-\bar{EA})^{2}-2\hat{\beta}(p_{i}-\bar{p})(EA_{i}-\bar{EA})}{N}\nonumber \\
= & \beta^{2}\lambda^{2}\sigma_{p}^{2}+\sigma_{EA}^{2}-2\beta\lambda\sigma_{p,EA},\label{eq:plim}
\end{align}

using the standard errors-in-variables formula $\textrm{plim}\hat{\beta}=\lambda\beta$,
and where $\sigma_{p,EA}$ is the covariance between $p$ and $EA$.

Write
\begin{align*}
\sigma_{p,EA} & =\textrm{Cov}(p^{*}+\eta,\alpha+\beta p^{*}+\varepsilon)\\
 & =0+\beta\sigma_{p^{*}}^{2}+\beta\textrm{Cov}(\eta,p^{*})+\textrm{Cov}(\eta,\varepsilon)\\
 & =\beta\sigma_{p^{*}}^{2}
\end{align*}

using the errors-in-variables assumption that $\eta$ is independent
of $p^{*}$ and $\varepsilon$. Plug this into (\ref{eq:plim}) to
give
\begin{align*}
 & \text{plim}\frac{\sum_{i=1}^{N}(\hat{\beta}(p_{i}-\bar{p})-(EA_{i}-\bar{EA}))^{2}}{N}\\
= & \beta^{2}\lambda^{2}\sigma_{p}^{2}+\sigma_{EA}^{2}-2\beta^{2}\lambda\sigma_{p^{*}}^{2}\\
= & \beta^{2}\lambda\sigma_{p^{*}}^{2}+\sigma_{EA}^{2}-2\beta^{2}\lambda\sigma_{p^{*}}^{2}\textrm{, using \ensuremath{\sigma_{p^{*}}^{2}=\lambda\sigma_{p}^{2}},}\\
= & \sigma_{EA}^{2}-\beta^{2}\lambda\sigma_{p^{*}}^{2}
\end{align*}

Inserting this into (\ref{eq:r-squared}) gives
\begin{align*}
\textrm{plim}\hat{R}^{2} & =1-\frac{\sigma_{EA}^{2}}{\sigma_{EA}^{2}}+\frac{\beta^{2}\lambda\sigma_{p^{*}}^{2}}{\sigma_{EA}^{2}}\\
 & =\lambda\frac{\beta^{2}\sigma_{p^{*}}^{2}}{\sigma_{EA}^{2}}\\
 & =\lambda R^{2}
\end{align*}
since for bivariate regression $R^{2}$ is the square of the correlation
between $EA$ and $p^{*}$, which can be written $r=$$\beta\frac{\sigma_{p^{*}}^{2}}{\sigma_{EA}\sigma_{p^{*}}}$.
We thank an anonymous commenter on https://stats.stackexchange.com/questions/238878/how-do-errors-in-variables-affect-the-r2
for much of this proof.

>>>>

Because selection effects are concentrated in lower-income groups, they may also
increase inequality with respect to polygenic scores. For example, Figure
\@ref(fig:plot-mean-EA3-by-income) graphs mean polygenic scores for educational
attainment (EA3) among children from households of different income groups. The
blue bars show the actual means, i.e. parents' mean polygenic score weighted by
number of children. The grey bars show the hypothetical means if all households
had equal numbers of children. Natural selection against genes associated with
educational attainment, which lowers the mean, is stronger at the bottom of the
income distribution, and this increases the differences between groups.
Overall, natural selection increases the correlation of polygenic scores with
income for `r n_increases` out of 33 polygenic scores, with a median
percentage increase of `r median_pct_change`% (Appendix Table
\@ref(tab:tbl-inequality-effects)).  If inequalities in polygenic scores are
important for understanding social structure and mobility 
[@belsky2018genetic; @Rimfeld_2018; @harden2021genetic], then these increases
are substantive. Similarly, since many polygenic scores are predictive of
disease risk, they could potentially increase health inequalities. In general,
the evolutionary history of anatomically modern humans is related to disease
risk [@benton2021influence]; understanding the role of contemporary natural
selection may aid in mapping the genetic architecture of current health
disparities.

@@
